---
title: "Course Project 1 - Reproducible Research"
author: "James Leather"
date: "November 11, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment Instructions
1. Code for reading in the dataset and/or processing the data
2. Histogram of the total number of steps taken each day
3. Mean and median number of steps taken each day
4. Time series plot of the average number of steps taken
5. The 5-minute interval that, on average, contains the maximum number of steps
6. Code to describe and show a strategy for imputing missing data
7. Histogram of the total number of steps taken each day after missing values are imputed
8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
9. All of the R code needed to reproduce the results (numbers, plots, etc.) in the report

## Step 1
## Code for loading the dataset and processing the data
```{r, echo=TRUE}
activity <- read.csv("activity.csv")
# load packages
library(ggplot2)
library(dplyr)
# some info about the variables
str(activity)
#transforming the date column into date format
activity$date <- as.Date(activity$date)
```

## Step 2
## Histogram of total steps per day
```{r, echo = TRUE}
#create and print steps per day
StepsPerDay <- with(activity, aggregate(steps, by = list(date), FUN = sum, na.rm = TRUE))
names(StepsPerDay) <- c("date", "steps")
StepsPerDay

#Histogram of total steps per day
g <- ggplot(StepsPerDay, aes(x = steps))
g+geom_histogram(boundary=0,binwidth = 2500, col="red", fill="blue") + ggtitle("Histogram of steps per day")+xlab("Steps")+ylab("Frequency")+theme(plot.title = element_text(face="bold", size = 12))+scale_x_continuous(breaks=seq(0,25000,2500))+scale_y_continuous(breaks = seq(0,18,2))
```

## Step 3
## Mean and median number of steps per day
```{r, echo=TRUE}
# Calculate the mean and median steps per day
mean(StepsPerDay$steps, na.rm = TRUE)
median(StepsPerDay$steps, na.rm = TRUE)
```

## Step 4
## Time series plot of average daily steps
```{r, echo = TRUE}
#Time series plot of average daily steps
StepsPerTime <- with(activity, aggregate(steps, by = list(interval), FUN = mean, na.rm = TRUE))
names(StepsPerTime) <- c("interval", "mean")
plot(StepsPerTime$interval, StepsPerTime$mean, type = "l", col="red", lwd = 2, xlab = "Interval", ylab = "Average steps", main = "Average number of steps per interval")
```

## Step 5
## 5 minute interval with max number of steps
```{r,echo=TRUE}
# 5 minute average with max number of steps
StepsPerTime[which.max(StepsPerTime$mean),]$interval
```

## Step 6
## Code to describe and show a strategy for imputing missing data
In this case I will impute the missing values to be the mean value
```{r, echo=TRUE}
# Count total number of NAs
sum(is.na(activity$steps))

# fill missing values
imputedSteps <- StepsPerTime$mean[match(activity$interval, StepsPerTime$interval)]
activity.imputed <- transform(activity, steps = ifelse(is.na(activity$steps), yes = imputedSteps, no = activity$steps))

# New data set
total.activity.imputed <- aggregate(steps ~ date, activity.imputed, sum)
colnames(total.activity.imputed) <- c("Date", "Steps")

#converting to a data frame
total.activity.imputed.df <- data.frame(total.activity.imputed)
```

## Step 7 
## Histogram of total number of steps after imputing the missing values
```{r, echo=TRUE}

#histogram for total steps per day after imputing missing values
g <- ggplot(total.activity.imputed.df, aes(Steps))
g+geom_histogram(boundary=0, binwidth=2500, col="darkblue", fill="lightblue")+ggtitle("Histogram of steps per day")+xlab("Steps")+ylab("Frequency")+theme(plot.title = element_text(face="bold", size=12))+scale_x_continuous(breaks=seq(0,25000,2500))+scale_y_continuous(breaks=seq(0,26,2))

#mean steps per day with imputed missing values
mean(total.activity.imputed$Steps)

#median steps per day with imputed missing values
median(total.activity.imputed$Steps)
```

## Step 8
## Panel plot comparing average steps per 5 minute interval across weekdays and weekends
```{r, echo=TRUE}
# Updating format of the dates
activity$date <- as.Date(strptime(activity$date, format="%Y-%m-%d"))

# Creating a function that distinguises weekdays from weekends
activity$dayType <- sapply(activity$date, function(x) {
  if(weekdays(x) == "Saturday" | weekdays(x) == "Sunday")
  {y <- "Weekend"}
  else {y <- "Weekday"}
  y
})

# Creating the data set that will be plotted
activityByDay <-  aggregate(steps ~ interval + dayType, activity, mean, na.rm = TRUE)

# Plotting using ggplot2
dayPlot <-  ggplot(activityByDay, aes(x = interval , y = steps, color = dayType)) + 
  geom_line() + ggtitle("Average Daily Steps by Day Type") + 
  xlab("Interval") + 
  ylab("Average Number of Steps") +
  facet_wrap(~dayType, ncol = 1, nrow=2) +
  scale_color_discrete(name = "Day Type") 
dayPlot
```